---
title: "The Effects of Weather on Urban Delay in New York City"
author: "Deric Pan, dericpan"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning=FALSE)
```

To begin this project: the following libraries are imported and the file is read in like so:
```{r}
library(boot)
library(tidyverse)
library(stringr)
library(lubridate)
library(chron)
library(reshape2)
buses = read.csv("Bus_Breakdown_and_Delays.csv", na.strings = c("", "NA"))
```

The data is then cleaned:
```{r}
buses = buses %>% filter(Boro %in% c("Manhattan", "Bronx", "Brooklyn", "Queens", "Staten Island")) %>% na.omit(How_Long_Delayed) %>% select(School_Year, Route_Number, Reason, Occurred_On, Boro, How_Long_Delayed)
buses = buses %>% filter(Reason == "Heavy Traffic")

buses = buses %>% mutate(Occurred_On = as.POSIXct(buses$Occurred_On, format = "%m/%d/%y %H:%M")) %>% mutate(x = as.POSIXct(buses$Occurred_On, format = "%m/%d/%y %H:%M")) %>% mutate(Occurred_On = strftime(Occurred_On, format = "%H:%M:%S")) %>% mutate(Occurred_on = as.POSIXct(buses$Occurred_On, format = "%H:%M:%S")) %>% select(School_Year, Route_Number, Occurred_On, Boro, How_Long_Delayed, x)

buses$Boro <- droplevels(buses$Boro)
buses$School_Year = droplevels(buses$School_Year, exclude = "2018-2019")


buses$How_Long_Delayed <- gsub("([0-9-]*)(.*)","\\1",buses$How_Long_Delayed)
buses$How_Long_Delayed <- gsub("0-15","15",buses$How_Long_Delayed)
buses$How_Long_Delayed <- gsub("16-30","30",buses$How_Long_Delayed)
buses$How_Long_Delayed <- gsub("31-45","45",buses$How_Long_Delayed)
buses$How_Long_Delayed <- gsub("46-60","60",buses$How_Long_Delayed)
buses <- buses[ !(buses$How_Long_Delayed %in% c(1,2,3)), ]

buses$How_Long_Delayed = as.numeric(buses$How_Long_Delayed)
buses = buses %>% na.omit(How_Long_Delayed)

buses = buses %>% mutate(Hour_Of_Day = hour(x))
buses <- buses %>% 
          mutate(Time_Of_Day = case_when(
                 .$Hour_Of_Day >= 6 & .$Hour_Of_Day < 10 ~ 'Morning',
                  .$Hour_Of_Day >= 10 & .$Hour_Of_Day < 16 ~ 'Midday',
                    TRUE ~ "Evening"
             ))
worked = buses %>% select(How_Long_Delayed, Time_Of_Day)
```

Data:
```{r}
ggplot(buses, aes(x=buses$How_Long_Delayed,fill = buses$Time_Of_Day)) +
  geom_histogram(binwidth = 10)+scale_x_continuous(breaks = c(15,30,45,60,75,90,105,120,135,150))+ggtitle("Bus Delay Length Count") + labs(title = "Bus Delay Counts", x = "Minutes", y = "Count")
```

```{r}
plot(density(rpois(100,20)))
```

```{r}
ggplot(buses, aes(x=buses$How_Long_Delayed,fill = buses$School_Year)) +
  geom_histogram(binwidth = 10)+scale_x_continuous(breaks = c(15,30,45,60,75,90,105,120,135,150)) + labs(title = "Bus Delay Counts by Year", x = "Minutes", y = "Count")
```

```{r}
boxplot(How_Long_Delayed ~ Time_Of_Day, data = buses, main = "Box Plot of Delays (minutes) by Time of Day", xlab = "Time Of Day", ylab = "Minutes")
```
Running Wilcox Tests:
```{r}
wilcox.test(How_Long_Delayed ~ Time_Of_Day, data = buses, subset = Time_Of_Day %in% c("Midday", "Morning"), conf.level = .95)

#p-value extremely high, so we know that we fail to reject the null hypothesis that they have equal medians

wilcox.test(How_Long_Delayed ~ Time_Of_Day, data = buses, subset = Time_Of_Day %in% c("Midday", "Evening"), conf.level = .95)

wilcox.test(How_Long_Delayed ~ Time_Of_Day, data = buses, subset = Time_Of_Day %in% c("Morning", "Evening"), conf.level = .95)
```

Running the code below produces a vector of p.values:
```{r}
y = buses$How_Long_Delayed
a = buses$How_Long_Delayed[buses$Time_Of_Day == "Morning"]
b = buses$How_Long_Delayed[buses$Time_Of_Day == "Evening"]

thetas = seq(-5,5, length.out = 10)

wp_vec = vector()
power_vec = vector()
for (theta in thetas) {
  new_a = a + theta
  wp = wilcox.test(new_a, b)$p.value
  wp_vec = append(wp_vec, wp)
}
```

Here, assume that mornings are distributed by poisson(20), simulate power of wilcox test with a+theta for morning = a & theta = 1.25, compare the two methods 
```{r}
wts.p = replicate(1000, {
  wts = wilcox.test(rpois(100, 20),
               rpois(100,20) + 1.25)
  wts$p.value <= 0.05 
  })
binom.test(sum(wts.p), 1000, conf.level = .95)

kts.p = replicate(1000, {
  kts = ks.test(rpois(100,20),
                rpois(100,20) + 1.25)
  kts$p.value <= 0.05
})
binom.test(sum(kts.p), 1000, conf.level = .95)


```
Create 95% confidence interval for thetas:
```{r}
range(thetas[wp_vec >= .05])
```
Test power with theta in 2.3
```{r}
wts.p = replicate(1000, {
  wts = wilcox.test(rpois(100, 20),
               rpois(100,20) + 2.3)
  wts$p.value <= 0.05 
  })
binom.test(sum(wts.p), 1000, conf.level = .95)
```
Test power with increasing thetas:
```{r}
powers_vec = vector()
conf_95_thetas = seq(.5556, 3.8889, length.out = 100)
for (theta in conf_95_thetas){
  wts.p = replicate(1000, {
  wts = wilcox.test(rpois(100, 20),
               rpois(100,20) + theta)
  wts$p.value <= 0.05 
  })
powers_vec = append(powers_vec,binom.test(sum(wts.p), 1000, conf.level = .95)$estimate)
}
```
Plot power curve:
```{r}
plot(conf_95_thetas, powers_vec, type = "l", xlab = "Theta", ylab = "Power")
```
Observe medians:
```{r}
median(buses$How_Long_Delayed[buses$Time_Of_Day == "Morning"])
median(buses$How_Long_Delayed[buses$Time_Of_Day == "Midday"])
median(buses$How_Long_Delayed[buses$Time_Of_Day == "Evening"])
```
Observe means:
```{r}
mean(buses$How_Long_Delayed[buses$Time_Of_Day == "Morning"])
mean(buses$How_Long_Delayed[buses$Time_Of_Day == "Midday"])
mean(buses$How_Long_Delayed[buses$Time_Of_Day == "Evening"])
```
Bootstraps for mean delay in minutes for respective times of day:
```{r}
mean_boot = function(data, index) {
  data_star = data[index,]
  mean_bus = mean(data_star$`How_Long_Delayed`[data_star$Time_Of_Day == "Morning"])
  return(mean_bus)
}  
boot_mean = boot(buses, statistic = mean_boot, R = 1000)
boot.ci(boot_mean, type = "norm")
```
```{r}
mean_boot = function(data, index) {
  data_star = data[index,]
  mean_bus = mean(data_star$`How_Long_Delayed`[data_star$Time_Of_Day == "Midday"])
  return(mean_bus)
}  
boot_mean = boot(buses, statistic = mean_boot, R = 1000)
boot.ci(boot_mean, type = "norm")
```
```{r}
mean_boot = function(data, index) {
  data_star = data[index,]
  mean_bus = mean(data_star$`How_Long_Delayed`[data_star$Time_Of_Day == "Evening"])
  return(mean_bus)
}  
boot_mean = boot(buses, statistic = mean_boot, R = 1000)
boot.ci(boot_mean, type = "norm")
```